Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-27 paper-1/manuscript.tex
Editing / mostly stats updates

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to dev target trials are larger due to lower $V_{TA}$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide median contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant correlation (two-sided t-test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Average $V_{TA}$ was lower in dev than msc (T = 1.24e+03, p = 2.07e-10, median = -0.074 mV; one-sided Wilcoxon, n = 120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

@@

Correlating the non-target response and thresholds excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev (t = -7.61, p = 3.5e-12, median = -0.044; one-sided t-test, n = 800) almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.994 and a consistent reduction in the average non-target response between msc and dev sequences (t = -11.6, p = 1.46e-21, median = -0.058; one-sided t-test, n = 120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $V_{TA}$ in dev is a consequence of lower dev non-target response magnitude.
\textbf{a} $V_{TA}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B, C, D, and E in msc). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation and the contrast in $V_{TA}$ at the start of target trials, and associated linear regression.
\textbf{d} Network-wide median contrast (dev - msc) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\Delta V_{TA}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided t-test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

@@

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and therefore decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (a) responds very quickly to stimulation in B, (b) responds more strongly to non-target stimuli in dev than msc, and (c) has increased TA levels as a result, reducing the later response in the same way as demonstrated in Figure \ref{fig4} in reverse.

To see this in the sample network, consider first the latency of responses to B across all conditions, shown by way of rank order in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward from the stimulus site. We did notice that the early response appears to gravitate towards the stimulus site for A, but confirmed that this was the case in all conditions separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B in any condition, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the response contrast (dev - msc) to the contrast in $V_{TA}$ at the start of B trials. Across all neurons, a strong linear relationship (Pearson's $\rho$ = 0.98) echoed the related finding in Figure \ref{fig5}c. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the dev sequence, while the network as a whole decreased. Correspondingly, the TA levels of this portion of the network also tended to increase, in clear contrast to the network as a whole.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $V_{TA}$ contrasts were robustly correlated (Median Pearson's $\rho$ 0.996, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon test for $\rho > 0$, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $V_{TA}$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in dev than msc (T = 6.98e+03, p = 8.48e-19; one-sided Wilcoxon, n = 120), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $V_{TA}$ followed the same pattern, with the early portion exhibiting stronger TA in dev (T = 4.63e+03, p = 2.21e-17), while the late portion was more recovered (T = 26, p = 2.97e-17).

@@

As noted previously, short-term depression alone exerted no deviance detection effect, but its addition did enhance the deviance detection effect established by TA. To understand the basis of this apparent synergy, we contrasted the TA only condition examined in the previous section against the same networks with both TA and STD. On average, adding STD caused a reduction in response magnitude across the board, as we would expect from a depression mechanism (data not shown). However, this reduction was not uniform: Target responses were reduced most in the std sequence, and least in the dev sequence.

@@

This pattern was reflected also in the aggregate across networks and stimuli. Firstly, the contrast (dev - msc) in the target response (Figure \ref{fig7}d, left) was positive in both models, but the full model contrast was greater than the ablated model contrast (t = 2.99, p = 0.00168; one-sided t-test, n = 120). Referring to the previous section, we identify the cause for this in an enhanced contrast in $V_{TA}$, as shown in Figure \ref{fig7}e; the contrast (dev - msc) in the full model was more negative than that in the ablated model (t = -3.49, p = 0.000342). Lastly, returning to Figure \ref{fig7}d (right), we see that the driver of $V_{TA}$, namely the non-target response size, also showed an enhanced contrast (dev - msc) in the full model relative to the ablation (t = -3.19, p = 0.000921). To sum up, the reason for the larger target dev response relative to msc, was the smaller non-target dev response relative to msc; furthermore, since the relationships established in the previous section remained largely unchanged, this effect appeared to be mediated primarily through TA.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to dev target trials are larger due to lower $V_{TA}$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in dev sequence and msc sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide median contrast (dev - msc) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (two-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in dev (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Median $V_{TA}$ was lower in dev than msc (T = 514, p = 1.67e-16, median = -0.094 mV; one-sided Wilcoxon, n = 120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

@@

Correlating the non-target response contrast (dev - msc) with the contrast in $V_{TA}$ across excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response between msc and dev sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $V_{TA}$ in dev is a consequence of lower dev non-target response magnitude.
\textbf{a} $V_{TA}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages for dev and msc sequences, respectively; right column, contrast between dev and msc averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the msc sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in dev, B, C, D, and E in msc). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between the contrast (dev - msc) in average number of spikes fired in response to non-target stimulation and the contrast in $V_{TA}$ at the start of target trials, and associated linear regression.
\textbf{d} Network-wide median contrast (dev - msc) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\Delta V_{TA}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

@@

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and thereby decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (a) responds very quickly to stimulation in B, (b) responds more strongly to non-target stimuli in dev than msc, and (c) has increased TA levels as a result, reducing the later response in the same way as demonstrated in Figure \ref{fig4} in reverse.

To see this in the sample network, consider first the latency of responses to B across all conditions, shown by way of rank order in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward from the stimulus site. We did notice that the early response appears to gravitate towards the stimulus site for A, but confirmed that this was the case in all conditions separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B in any condition, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the response contrast (dev - msc) to the contrast in $V_{TA}$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993) echoed the related finding in Figure \ref{fig5}c. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the dev sequence, while the response of the network as a whole decreased. Correspondingly, the TA levels of this portion of the network also tended to increase, in clear contrast to the network as a whole.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $V_{TA}$ contrasts were robustly correlated (Median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon test for $\rho > 0$, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $V_{TA}$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in dev than msc (T = 6.98e+03, p = 8.48e-19; one-sided Wilcoxon, n = 120), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $V_{TA}$ followed the same pattern, with the early portion exhibiting stronger TA in dev (T = 4.63e+03, p = 2.21e-17), while the late portion was more recovered (T = 26, p = 2.97e-17).

@@

% TODO rev Fig7
As noted previously, short-term depression alone exerted no deviance detection effect, but its addition did enhance the deviance detection effect established by TA. To understand the basis of this apparent synergy, we contrasted the TA only condition examined in the previous section against the same networks with both TA and STD. On average, adding STD caused a reduction in response magnitude across the board, as we would expect from a depression mechanism (data not shown). However, this reduction was not uniform: Target responses were reduced most in the std sequence, and least in the dev sequence.

@@

This pattern was reflected also in the aggregate across networks and stimuli. Firstly, the contrast (dev - msc) in the target response (Figure \ref{fig7}d, left) was positive in both models, but the full model contrast was greater than the ablated model contrast (t = 2.99, p = 0.00168; one-sided t-test, n = 120). Referring to the previous section, we identify the cause for this in an enhanced contrast in $V_{TA}$, as shown in Figure \ref{fig7}e; the contrast (dev - msc) in the full model was more negative than that in the ablated model (t = -3.49, p = 0.000342). Lastly, returning to Figure \ref{fig7}d (right), we see that the driver of $V_{TA}$, namely the non-target response size, also showed an enhanced contrast (dev - msc) in the full model relative to the ablation (t = -3.19, p = 0.000921). To sum up, the reason for the larger target dev response relative to msc, was the smaller non-target dev response relative to msc; furthermore, since the relationships established in the previous section remained largely unchanged, this effect appeared to be mediated primarily through TA.
% end TODO
